Tempo and mode of promoter evolution as observed in the \textit{Paramecium aurelia} species complex

### 2019-04-13 manuscript.tex
Updated preliminary title
Added stashed manuscript.txt file

\title{Mode and tempo of promoter evolution as observed in the \textit{Paramecium aurelia} species complex}

@@

A major challenge in genomics is to decipher the precise instructions that regulate gene expression. The regulation of gene expression underpins fundamental processes within the cell, such as growth, development, the maintenance of homeostasis and metabolism. Transcription, a major step in gene expression, is controlled through proximal, \textit{cis}-regulatory regions bound by \textit{trans}-acting transcription factors. Investigations across select eukaryotes have drawn attention to the important role of \textit{cis}-regulatory sequences in the evolution of gene expression differences \citep{Wittkopp:2008ki, Wittkopp:2011bc, Siepel:2014hd}. Work over the previous two decades have emphasized the contribution of \textit{cis}-regulatory sequence differences toward changes in gene expression \citep{Wittkopp:2004cy}. The fulcrum and first major step of gene expression is transcription initiation, which in eukaryotes begins with the mediated interaction of RNA polymerase II complex with the promoter\citep{Kadonaga:2011gz}. The locations of eukaryotic \textit{cis}-regulatory sequences, including promoters, cannot be predicted from genomes with precision.\\
Comparative and functional genomic approaches have shed light on the contributions of \textit{cis}-regulatory sequences to phenotypic evolution and species divergence. Investigating F1 hybrids of \textit{D. melanogaster} and \textit{D. simulans}, Wittkopp and colleagues \citep{Wittkopp:2004cy} showed that interspecific expression differences were primarily caused by changes acting in \textit{cis}, not \textit{trans}. Evolutionary changes in gene expression patterns have been noted in a number of instances, including in expression timing (\textit{i.e.} heterochrony) \citep{Wray:1989ku}, expression levels (\citep{Crawford:1999ic}), spatial differences \citep{Abzhanov:2000vb}, and sex-specific expression \citep{Kopp:2000bl}; reviewed in \citep{Wray:2003kn}. The structure and conservation of the regulatory regions of select genes, including the hox \citep{Kmita:2002hg} and pax \citep{Plaza:1999kc} gene complexes. The advent of large-scale functional genomics methods has identified sequence variants within regulatory regions that contribute to expression differences \citep{Lappalainen:2013el,Pickrell:2010bq, Schor:2017fw}; a number of these have been functionally characterized. These efforts have revealed the cellular mechanisms of transcriptome and regulatory variation, they have largely been confined to major model systems, especially human, and have not focused on exploring the changes to \textit{cis}-regulatory regions over evolutionary time-scales. As it stands, the precise \textit{cis}-regulatory sequence differences that accompany---or indeed underpin---species divergence remain largely obscure. One plausible explanation for this is the difficulty in predicting \textit{cis}-regulatory regions from genomic sequence alone. Accurate estimation of promoter positions requires empirical, usually functional genomic evidence. At present, the best high-throughput approach to identify promoters at genome-scale is transcription start site (TSS) profiling, which includes CAGE \citep{Murata:2014gpb} and RAMPAGE \citep{Batut:2013fu}, among others. While TSS profiling methods differ in technical between themselves, these protocols all capture the 5$`$-ends of capped mRNAs, sequence their corresponding cDNAs and align the reads to the genome to identify the TSSs present within a given transcriptome. Clustering gene-adjacent TSSs defines a promoter, transcription start region (TSR), at single base-pair resolution, thereby providing genomic locations for \textit{cis}-regulatory regions within a species \citep{Lenhard:2012ena,Rach:2009ct}. TSS profling studies in a variety of model organisms has demonstrated that the shape of TSS distributions (\textit{i.e.} promoter shape) at promoters is related to the promoter class and correlates with the function of the associated gene \citep{Carninci:2006in, Rach:2009ct,Hoskins:2011io,Raborn:2016cr}. In addition, recent work done in \textit{Drosophila melanogaster} indicates that promoter shape is itself a quantitative trait \citep{Schor:2017fw}, a finding that carries valuable implications for the understanding of the evolution of \textit{cis}-regulatory regions. \\ 

>>>>

\title{Tempo and mode of promoter evolution as observed in the \textit{Paramecium aurelia} species complex}

@@

A major challenge in genomics is to decipher the precise instructions that regulate gene expression. The regulation of gene expression underpins fundamental processes within the cell, such as growth, development, the maintenance of homeostasis and metabolism. Transcription, a major step in gene expression, is controlled through \textit{cis}-regulatory regions bound by \textit{trans}-acting transcription factors. Investigations across select eukaryotes have drawn attention to the important role of \textit{cis}-regulatory sequences in the evolution of gene expression differences \citep{Wittkopp:2008ki, Wittkopp:2011bc, Siepel:2014hd}. Work over the previous two decades have emphasized the contribution of \textit{cis}-regulatory sequence differences toward changes in gene expression \citep{Wittkopp:2004cy}. The fulcrum and first major step of gene expression is transcription initiation, which in eukaryotes begins with the mediated interaction of RNA polymerase II complex with the promoter, a \textit{cis}-regulatory region located proximal to the gene \citep{Kadonaga:2011gz}. The locations of eukaryotic \textit{cis}-regulatory sequences, including promoters, cannot be predicted from genomes with precision.\\
Comparative and functional genomic approaches have shed light on the contributions of \textit{cis}-regulatory sequences to phenotypic evolution and species divergence. Investigating F1 hybrids of \textit{D. melanogaster} and \textit{D. simulans}, Wittkopp and colleagues \citep{Wittkopp:2004cy} showed that interspecific expression differences were primarily caused by changes acting in \textit{cis}, not \textit{trans}. Evolutionary changes in gene expression patterns have been noted in a number of instances, including in expression timing (\textit{i.e.} heterochrony) \citep{Wray:1989ku}, expression levels (\citep{Crawford:1999ic}), spatial differences \citep{Abzhanov:2000vb}, and sex-specific expression \citep{Kopp:2000bl}; reviewed in \citep{Wray:2003kn}. The structure and conservation of the regulatory regions of select genes, including the hox \citep{Kmita:2002hg} and pax \citep{Plaza:1999kc} gene complexes, were established. More recently, the advent of large-scale functional genomics methods has identified sequence variants within regulatory regions that contribute to expression differences \citep{Lappalainen:2013el,Pickrell:2010bq, Schor:2017fw}; a number of these have been functionally characterized. Global functional genomics efforts have revealed the cellular mechanisms of transcriptome and regulatory variation, but have largely been confined to major model systems, especially human, and have not focused on exploring the changes to \textit{cis}-regulatory regions over evolutionary time-scales. As it stands, the precise \textit{cis}-regulatory sequence differences that accompany---or indeed underpin---species divergence remain largely obscure. Another potential explanation for this is the difficulty in predicting \textit{cis}-regulatory regions from genomic sequence alone. Accurate estimation of promoter positions requires empirical evidence. At present, the best high-throughput approach to identify promoters at genome-scale is transcription start site (TSS) profiling, which includes CAGE \citep{Murata:2014gpb} and RAMPAGE \citep{Batut:2013fu}, among others. While TSS profiling methods differ in technical between themselves, these protocols all capture the 5$`$-ends of capped mRNAs, sequence their corresponding cDNAs and align the reads to the genome to identify the TSSs present within a given transcriptome. Clustering gene-adjacent TSSs defines a promoter, transcription start region (TSR), at single base-pair resolution, thereby providing genomic locations for \textit{cis}-regulatory regions within a species \citep{Lenhard:2012ena,Rach:2009ct}. TSS profling studies in a variety of model organisms has demonstrated that the shape of TSS distributions (\textit{i.e.} promoter shape) at promoters is related to the promoter class and correlates with the function of the associated gene \citep{Carninci:2006in, Rach:2009ct,Hoskins:2011io,Raborn:2016cr}. In addition, recent work done in \textit{Drosophila melanogaster} indicates that promoter shape is itself a quantitative trait \citep{Schor:2017fw}, a finding that carries valuable implications for the understanding of the evolution of \textit{cis}-regulatory regions. \\
